### 2025-01-20 paper/icml25/first/sects/background.tex
revised bg

@@

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, F, R, T, \gamma, \mu)$, where $\mathcal{S}$ represents the state space, $\mathcal{A}$ denotes the action space, $F:\mathcal{S} \times \mathcal{A} \rightarrow \mathcal{S}$ is the transition function, $R:\mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, $\gamma \in [0, 1]$ is the discount factor, and $T \in \mathbb{N}$ is the time horizon. The initial state distribution is denoted by $\mu$.

Further, let $\mathcal{T}$ denote the space of trajectories, defined as $\mathcal{T} = (\mathcal{S} \times \mathcal{A} \times \mathcal{S} \times \mathbb{R})^*$, where each trajectory consists of a sequence of starting states, actions, subsequent states, and rewards. We will use the symbols $\tau=((s_0, a_0, s_1, r_0), (s_1, a_1, s_2, r_1), \ldots)$ to denote a generic trajectory in $\mathcal{T}$. When dealing with multiple trajectories, we will use the notation $s_i^\tau, a_i^\tau$, and $r_i^\tau$.

A policy is a function $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maps states to actions. We can use the policy to sample trajectories from the MDP by iteratively applying the transition and reward function and the policy. In this case, we have that the $i$-th element of a trajectory is $(s_i, a_i, s_{i+1}, r_i) = (s_i, \pi(s_i), F(s_i, a_i), R(s_i,a_i,F(s_i, a_i))$. The initial state $s_0$ is sampled from the initial state distribution $\mu$.

For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled by following the policy $\pi$.

The goal of the MDP is to find a policy that maximizes the expected return 
$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r_t^\tau \right]$$

@@

In this work, we will often approximate functions, such as transition ($F$) or reward ($R$), with neural network. Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$. For instance, the transition and its corresponding neural network are denoted as $F$ and $F_{\theta}$, respectively. The corresponding loss will be denoted as $\mathcal{L}_\theta^g$ and it will aim to minimize the differences between $g$ and $g_\theta$.

While we use a unique set of parameters $\theta$ for all the networks. Each network can either use its own subset of $\theta$ or it can share parameters with other networks. In practice, the parameters will never be shared.

@@

To solve an MDP $\mathcal{M}$, SHAC trains a policy $\pi_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$ for a batch of $N$ trajectories $B = \{\tau_1, \ldots, \tau_N\}$ sampled from the MDP. The loss is defined as:  
\begin{equation}\label{eq:shac}
    \mathcal{L}_\theta^{\pi}(B) = \frac{1}{NH}\sum_{\tau\in B} \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^\tau, a_t^\tau) + \gamma^H V_\theta(s^\tau_{t_0+T})
\end{equation}
where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$. From Equation~\ref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$. While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

The value function $V_\theta$ is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $V$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{V}:\mathcal{T}^N\rightarrow\mathbb{R}$.

$$ \mathcal{L}_\theta^V(B)=\frac{1}{NT}\sum_{\tau\in B}\sum_{t=0}^T\left|\left| V_\theta(s^\tau_t) - V(s^\tau_t) \right|\right|^2 $$
$$ V(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
$$ G_t^k = \left(\sum_{l=0}^{k-1}\gamma^l r_{t+l}\right) + \gamma^k V(s_{t+k})$$

@@

Finally, Figure~\ref{fig:shac} presents an unrolled view of the SHAC algorithm.

>>>>

\newcommand{\agents}{\mathcal{N}}

@@

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: 
$$(\agents,\states,\{\observations{i}\}_{\scriptscriptstyle i\in\agents},\{\actions{i}\}_{\scriptscriptstyle i\in\agents},\{\stateobs{i}\}_{\scriptscriptstyle i\in\agents}, \{\rewards{i}\}_{\scriptscriptstyle i\in\agents}, \transition, \horizon, \discount, \initdist)$$
Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\observations{i}$ and $\actions{i}$ are the observation space and action space for the $i$-th agent ($i\in\agents$). $\stateobs{i}:\states\rightarrow\observations{i}$ is the projection to the observation space and $\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}$ is the reward function for the $i$-th agent. $\transition:\states\times\{\actions{i}\}_{i\in\agents}$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\{\actions{i}\}_{i\in\agents}\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent. Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space. We will denote the collective policy, determined by the collection of agent-specific policy simply with $\pi$. We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions. In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$. The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$. Finally, $s_0$ is sampled from the initial state distribution $\initdist$. For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.

The goal of the MDP is to find a policy that maximizes the expected return 
$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]$$

@@

In this work, we will often approximate functions (such as transition $\transition$) with neural network. Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$. For instance, the transition and its corresponding neural network are denoted as $F$ and $F_{\theta}$, respectively. The corresponding loss will be denoted as $\mathcal{L}_\theta^g$ and it will aim to minimize the differences between $g$ and $g_\theta$.

While we use a unique set of parameters $\theta$ for all the networks. Each network can either use its own subset of $\theta$ or it can share parameters with other networks. In practice, the parameters will be shared only for the policies.

For simplicity, we formally treat only the case where the policy observe the full state to make the action, meaning that $\stateobs{i}$ is the identity function. However, in our experiments, we consider also scenarios where the full state is not available to the policies.

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

The value function $\valuef_\theta$ is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $\valuef$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{\valuef}:\trajectories^\bs\rightarrow\mathbb{R}$.

$$ \mathcal{L}_\theta^\valuef(\batch)=\frac{1}{\bs\mh}\sum_{\tau\in\batch}\sum_{t=0}^\mh\left|\left| \valuef_\theta(s^\tau_t) - \valuef(s^\tau_t) \right|\right|^2 $$
$$ \valuef(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
$$ G_t^k = \left(\sum_{l=t}^{k-1+t}\sum_{i\in\agents}\gamma^{l-t} r_{li}\right) + \gamma^k V(s_{t+k})$$

@@

Finally, \Cref{fig:shac} presents an unrolled view of the SHAC algorithm for a single-agent scenario.

### 2025-01-20 paper/icml25/first/sects/shacpp.tex
revised bg

As mentioned previously, we aim to address the limitations of SHAC by approximating the gradients of a differentiable environment ($R$ and $F$) with neural networks ($R_\theta$ and $F_\theta$). In particular, we aim to train a neural network to approximate the reward function by minimizing: 
$$ \mathbb{E}_{\tau\sim\pi}\left[\left|\left| \frac{1}{T}\sum_{i=0}^{T-1} R_{\theta}(s_i^\tau, a_i^\tau, s_{i+1}^\tau) - R(s_i^\tau, a_i^\tau, s_{i+1}^\tau) \right|\right|^2\right] $$

In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment. The neural network is then trained to minimize the mean squared error (MSE) loss. Specifically, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \mathcal{L}_\theta^R(B) = \frac{1}{N}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - r \right|\right|^2 $$
Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $H$ actions from $\mathcal{A}^H$, predicts the next $H$ observations from $\mathcal{S}^H$. We aim to minimize:

\begin{equation}
\begin{aligned}
\mathbb{E}_{\tau\sim\pi}\left[\frac{1}{NH}\sum_{t=0}^{T-H-1}\sum_{i=0}^{H-1} \left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - \right.\right.\right. \\ \left.\left.\left. s_{t+i+1}^\tau \right|\right|^2\right] 
\end{aligned}
\end{equation}

In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{NH}\sum_{\tau \in B}\sum_{t=t_0}^{t_0+H-1}\left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - s_{t+1}^\tau \right|\right|^2 $$
Where $B$ represents a batch of trajectories with horizon $H$ sampled uniformly from the cache.

Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space. That is, suppose the action space is $(a,b)$ (with $a<b$), we minimize the loss $\mathcal{L'}_\theta^\pi:$
\begin{figure*}[!t]
    \centering
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
\end{figure*}

$$\mathcal{L}_\theta^\pi(B) + \alpha \frac{1}{NH}\sum_{\tau\in B}\sum_{t=t_0}^{t_0+H-1} d(a_t^\tau,(a,b))$$
$$d(x,(a,b)) = \begin{cases}(a-x)^2 & \text{if } x < a \\ (x-b)^2 & \text{if } x > b \\ 0 & \text{otherwise} \end{cases}$$

>>>>

As mentioned previously, we aim to address the limitations of SHAC by approximating the gradients of a differentiable environment ($R$ and $F$) with neural networks ($R_\theta$ and $F_\theta$). In particular, we aim to train a neural network to approximate the reward function by minimizing: 
\begin{align*}
    \mathbb{E}_{\tau\sim\pi}\left[\left\vert\left\vert \frac{1}{\horizon\na}\sum_{j=0}^{\horizon-1} \sum_{i\in\agents} {\rewards{i}}_\theta(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right.\right.\right. -\\ \left.\left.\left. \rewards{i}(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right\vert\right\vert^2\right] 
\end{align*}

In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment. The neural network is then trained to minimize the mean squared error (MSE) loss. Specifically, we minimize the loss $\mathcal{L}_\theta^{\rewards{i}}:(\states\times\actions{i}\times\states)^N\rightarrow\mathbb{R}$:
$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,r \in \batch}\left|\left| {\rewards{i}}_\theta(s, a) - r \right|\right|^2 $$
Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $(\{\actions{i}\}_{i\in\agents})^\mh$, predicts the next $\mh$ observations from $\states^\mh$. We aim to minimize:
%\sum_{j=0}^{\scriptscriptstyle\mh-1}
\begin{align*}
    \mathbb{E}_{\tau\sim\pi}\left[\frac{1}{\bs\mh}\sum_{t=0}^{\scriptscriptstyle\horizon-\mh-1}\left|\left| \transition_{\theta}(s_t^\tau, \{a^\tau_t\}_{i}^{\na},\dots,\{a^\tau_{\scriptscriptstyle t+H-1}\}_{i}^{\na}) \right.\right.\right.-\\\left.\left.\left. \left[s_t^\tau,\dots,s_{t+h}^\tau\right] \right|\right|^2\right] 
\end{align*}

In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, \{a_1^\tau\}_i^\na, \dots, \{a_{\mh}^\tau\}_i^\na) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.

Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space. Suppose the action space is $(l,r)$ (with $l<r$) for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$
\begin{figure*}[!t]
    \centering
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion.png}
        \caption{Dispersion}
        \label{fig:dispersion}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/transport.png}
        \caption{Transport}
        \label{fig:transport}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/discovery.png}
        \caption{Discovery}
        \label{fig:discovery}
    \end{subfigure}
    \begin{subfigure}[b]{0.24\textwidth}
        \includegraphics[width=\textwidth]{figs/sampling.png}
        \caption{Sampling}
        \label{fig:sampling}
    \end{subfigure}
    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
\end{figure*}

$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(a,b))$$
$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > e \\ 0 & \text{otherwise} \end{cases}$$
